---
title: "Distributions"
output: pdf_document
---

# 1. Introduction

This is the R track of the *Distributions* workshop prepared by
Los Angeles County, ISAB.  Section numbering is intended to be
consistently referenced from the main workshop document available
at

<https://github.com/lacounty-isab/workshops/tree/master/distributions>

## 1.1. Preparation

Since R is oriented toward statistics, no special preparation is
required to work with distributions.
You can obtain a list of *baked-in* distributions by entering

```{r, eval=FALSE, echo=TRUE}
?Distributions
```

These are all part of the **stats** package which is available
in every R installation.

## 1.2 Conventions

R has a consistent naming convention for functions that work
with distributions - a single letter followed by the name of
the distribution.  The four single letters are

* `d` - density function
* `p` - percent point function (CDF)
* `q` - inverse of CDF
* `r` - random sampler

If we take the binomial distribution as an example, then
`dbinom` is a *binomial density function*.

```{r dbinom1}
dbinom(4, 10, 0.3)
```

This gives the probability of getting 4 successes after 10
attempts where each success has a probability of 0.3;
which is about `r round(dbinom(4,10,0.3), 3)*100` percent.