---
title: "R Regex API"
author: "ISAB Data Science Forum"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Regular Expressions
There are several regular expression functions in R defined in the
**base** package.  The most recognizable one is `grep`.
It takes

1. a pattern
2. a character vector (or something coercible to one)
3. options

and returns an vector of indices that matched.

```{r eg01}
lastNames <- c("Smith", "Jones", "Thomson", "Johnson")
sonNames <- grep("son", lastNames)
sonNames
```

## R `grep` results

The index can be used to select the matching items from the
original vector.

```{r eg02}
lastNames[sonNames]
```

Or the items that did **not** match.

```{r eg03}
lastNames[-sonNames]
```

## Grep Options: `value`

By default, `value=FALSE`.  This means return the index
of each match.  We saw in the last slide how we can use the
match indices to retrieve the original values.  This can
be done in one step by passing `value=TRUE`.

```{r value1}
grep("son", lastNames, value=TRUE)
```

## Other Grep Options

The following boolean options to `grep` default to `FALSE`.
They can be activated by setting them to `TRUE`.

* `ignore.case` - case insensitive match
* `invert` - return the non-matched results
* `perl` - use PCRE (Perl Compatible Regular Expression) syntax
* `fixed` - treat the pattern as a plain string (no fancy regular
   expresion interpretation)

```{r eg04}
noJo <- grep("Jo", lastNames, invert=TRUE, ignore.case=TRUE)
noJo
```

## Logical Indices: `grepl`

This is the "grep logical" function.
Instead of returning an list of integer indices that matched,
you can have a logical vector of the same length as the input vector.
`TRUE` values match; values do not.

```{r eg05}
hasSon <- grepl("son", lastNames)
hasSon
```
How many matches did we have?
```{r eg06}
table(hasSon)
```

## Substitution

The `sub` function performs substitution on the first match
for each element of the vector. 
Use the global version, `gsub`, to substitute for all matches
on each element of a vector.

```{r eg07}
sub("John", "David", lastNames)
```

Changed "Johnson" to "Davidson".  Left non-matching entries alone.

## Character Position

The `regexpr` function returns an vector of integers that represent
the offset of the first match for each element of the original vector.

```{r eg08, echo=FALSE}
lastNames
```
```{r eg09}
m <- regexpr("o\\w+son", lastNames)
m[1:4]
```

The return value also maintains a vector stored in an attribute
named `match.lengths`.  It represents the length of each match.

```{r eg10}
attr(m, "match.length")
```

## Drawbacks

- There doesn't seem to be support for regular expression groupings.

- Processing a file one line at a time is tedious.  It's possible to
  read and process a single line at a time.  But this is **very**
  slow compared to the same task in Python (4 seconds vs 100 seconds).